Microxanox: an R package for simulating an aquatic \(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.

### 2023-09-08 inst/manuscript/revision_2/Response_rev2.md

The second document is the revised and clean manuscript, and the third is the revised manuscript with changes shown.

Below, we cover each of the latest reviewer comments in turn.

> It was difficult to follow the [Responses and Suggestions] in the new version. Those sections/paragraphs that have been rewritten, added, or modified in the paper should be included in the reply letter and explicitly highlight the parts that correspond precisely to each remark. The highlighted paper and the clean version should be shown at the end of the reply letter.

Apologies that we provided materials that were difficult to follow. With the current revision we provide two versions of the revised manuscript, as requested, one a clean version, and one with changes highlight. We also, as requested, include all revised text in this response document and then reference the location in the revised document by line number. *Rainer, please check that this is so, before resubmission. Wait for our final version of the text before inserting the correct line numbers below.*

### 2023-09-08 inst/manuscript/revision_2/manuscript_microxanox_rev2.Rmd
Changes concerning term inology

To make the simulation run with multiple species per functional group, we expressed different species' characteristics in the elements of vectors and matrices. We also coded the ordinary differential equations to include the vectors and matrices and used matrix mathematics. In this way, we made it possible to run simulations with different numbers of species without having to change the underlying code.

@@

A simulation is run using the `run_simulation()` function. In this function, the ODEs are solved using the function `ode()` in the package *deSolve* package [@Soetaert2010]. The `run_simulation()` function needs only one argument - an object as created by the function `new_runsim_parameter()`. The parameter object returned by `new_runsim_parameter()` contains, among other things, the `strain_parameter` object, which can be created by the function `new_strain_parameter()`. For a detailed description of the parameter objects, their meaning and how they are created and have values set and changed, please see the *User Guide* which accompanies the package or is available at [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html).

@@

One approach to finding how the ecosystem responds to environmental driver change is to examine the relationship between the environmental driver value and the numerical value of a system's state. The first component of this approach is to run the simulation for a long time and record the final state (i.e., the state of the system at the end of a simulation). When one does this across a range of environmental conditions, one can discovers how the final state of the system changes with the environmental conditions. The package contains functionality for finding final states that correspond to values of one environmental driver, namely the value of oxygen diffusivity. In terms of non-linear systems analysis, this would be termed a *parameteric sensitivity analysis* being conducted by running an open loop of the dynamic system under a set of initial conditions and a subset of parameter values (here oxygen diffusivity).

When one wishes to be able to make conclusions about how the *steady state* (or an equilibrium point) of the system is affected by the environmental driver, it is very important to note that the *final state* (provided by the simulation) is not guaranteed to be a *steady state*. The software does not provide the user with a steady state. In order to somewhat safely assume that the final state is a steady state, the user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappeared, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper @Limberger2023, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation. Furthermore, the package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique.

Two numerical strategies for finding final states and their sensitivity to parameters are implemented. Two strategies are implemented in order to allow users to compare of their results. The first method runs a independent simulation for each combination of initial conditions and oxygen diffusivity (we term this the *Replication method*). This is the method used in the @Bush2017 study and was used to obtain the results in figures 3 and 4 of that article.

@@

The open source implementation and extension of the model used in @Bush2017 provides the means of reproducing the results published while at the same time provides the means of making unique, innovative, and important investigations of how ecosystems change as the environmental changes, and how biodiversity may modulate this.

The design of the package code and functionality is with reproducibility in mind (for other example of reproducibility directed software, see e.g. @Correndo2023, @Spillner2022 or @Fox2020): the combination of all parameters being in a single parameter object and the return of the simulation as a result object which inherits from the parameter object, provides a relatively easy-to-use framework to implement reproducible experiments.

>>>>

To make the simulation run with multiple species per functional group, we expressed different species' characteristics in the elements of vectors and matrices. We also coded the ordinary differential equations in matrix form to use matrix mathematics. In this way, we made it possible to run simulations with different numbers of species without having to change the underlying code.

@@

A simulation is run using the `run_simulation()` function. In this function, the ODEs are solved using the function `ode()` in the package *deSolve* [@Soetaert2010]. The `run_simulation()` function needs only one argument - an object as created by the function `new_runsim_parameter()`. The parameter object returned by `new_runsim_parameter()` contains, among other things, the `strain_parameter` object, which can be created by the function `new_strain_parameter()`. For a detailed description of the parameter objects, their meaning and how they are created and have values set and changed, please see the *User Guide* which accompanies the package or is available at [User Guide](https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html).

@@

One approach to finding how the ecosystem responds to environmental driver change is to examine the relationship between the environmental driver value and the numerical value of a system's state. The first component of this approach is to run the simulation for a long time and record the final state (i.e., the state of the system at the end of a simulation). When one does this across a range of environmental conditions, one discovers how the final state of the system changes with the environmental conditions. The package contains functionality for finding final states that correspond to values of one environmental driver, namely the value of oxygen diffusivity. In terms of non-linear systems analysis, this would be termed a *parametric sensitivity analysis* being conducted by running an open loop of the dynamic system under a set of initial conditions and a subset of parameter values (here oxygen diffusivity).

When one wishes to be able to make conclusions about how the *steady state* (or an equilibrium point) of the system is affected by the environmental driver, it is very important to note that the *final state* (provided by the simulation) is not guaranteed to be a *steady state*. The software does not provide the user with a steady state. In order to somewhat safely assume that the final state is a steady state, the user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappear, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper @Limberger2023, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation. Furthermore, the package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique.

Two numerical strategies for finding final states and their sensitivity to parameters are implemented. The first method runs a independent simulation for each combination of initial conditions and oxygen diffusivity (we term this the *Replication method*). This is the method used in the @Bush2017 study and was used to obtain the results in figures 3 and 4 of that article.

@@

The open source implementation and extension of the model used in @Bush2017 provides the means of reproducing the results published while at the same time provides the means of making unique, innovative, and important investigations of how ecosystems change as the environment changes, and how biodiversity may modulate this.

The design of the package code and functionality is with reproducibility in mind (for other example of reproducibility directed software, see e.g. @Correndo2023, @Spillner2022 or @Fox2020): the combination of all parameters being in a single parameter object and the return of the simulation as a result object which inherits the parameter values from the parameter object, provides a relatively easy-to-use framework to implement reproducible experiments.

### 2023-09-08 inst/manuscript/revision_2/manuscript_microxanox_rev2.tex
Changes concerning term inology

A simulation is run using the \texttt{run\_simulation()} function. In this function, the ODEs are solved using the function \texttt{ode()} in the package \emph{deSolve} package \citep{Soetaert2010}. The \texttt{run\_simulation()} function needs only one argument - an object as created by the function \texttt{new\_runsim\_parameter()}. The parameter object returned by \texttt{new\_runsim\_parameter()} contains, among other things, the \texttt{strain\_parameter} object, which can be created by the function \texttt{new\_strain\_parameter()}. For a detailed description of the parameter objects, their meaning and how they are created and have values set and changed, please see the \emph{User Guide} which accompanies the package or is available at \href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User Guide}.

@@

One approach to finding how the ecosystem responds to environmental driver change is to examine the relationship between the environmental driver value and the numerical value of a system's state. The first component of this approach is to run the simulation for a long time and record the final state (i.e., the state of the system at the end of a simulation). When one does this across a range of environmental conditions, one can discovers how the final state of the system changes with the environmental conditions. The package contains functionality for finding final states that correspond to values of one environmental driver, namely the value of oxygen diffusivity. In terms of non-linear systems analysis, this would be termed a \emph{parameteric sensitivity analysis} being conducted by running an open loop of the dynamic system under a set of initial conditions and a subset of parameter values (here oxygen diffusivity).

When one wishes to be able to make conclusions about how the \emph{steady state} (or an equilibrium point) of the system is affected by the environmental driver, it is very important to note that the \emph{final state} (provided by the simulation) is not guaranteed to be a \emph{steady state}. The software does not provide the user with a steady state. In order to somewhat safely assume that the final state is a steady state, the user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappeared, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper \citet{Limberger2023}, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation. Furthermore, the package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique.

Two numerical strategies for finding final states and their sensitivity to parameters are implemented. Two strategies are implemented in order to allow users to compare of their results. The first method runs a independent simulation for each combination of initial conditions and oxygen diffusivity (we term this the \emph{Replication method}). This is the method used in the \citet{Bush2017} study and was used to obtain the results in figures 3 and 4 of that article.

@@

The open source implementation and extension of the model used in \citet{Bush2017} provides the means of reproducing the results published while at the same time provides the means of making unique, innovative, and important investigations of how ecosystems change as the environmental changes, and how biodiversity may modulate this.

The design of the package code and functionality is with reproducibility in mind (for other example of reproducibility directed software, see e.g. \citet{Correndo2023}, \citet{Spillner2022} or \citet{Fox2020}): the combination of all parameters being in a single parameter object and the return of the simulation as a result object which inherits from the parameter object, provides a relatively easy-to-use framework to implement reproducible experiments.

>>>>

A simulation is run using the \texttt{run\_simulation()} function. In this function, the ODEs are solved using the function \texttt{ode()} in the package \emph{deSolve} \citep{Soetaert2010}. The \texttt{run\_simulation()} function needs only one argument - an object as created by the function \texttt{new\_runsim\_parameter()}. The parameter object returned by \texttt{new\_runsim\_parameter()} contains, among other things, the \texttt{strain\_parameter} object, which can be created by the function \texttt{new\_strain\_parameter()}. For a detailed description of the parameter objects, their meaning and how they are created and have values set and changed, please see the \emph{User Guide} which accompanies the package or is available at \href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User Guide}.

@@

One approach to finding how the ecosystem responds to environmental driver change is to examine the relationship between the environmental driver value and the numerical value of a system's state. The first component of this approach is to run the simulation for a long time and record the final state (i.e., the state of the system at the end of a simulation). When one does this across a range of environmental conditions, one discovers how the final state of the system changes with the environmental conditions. The package contains functionality for finding final states that correspond to values of one environmental driver, namely the value of oxygen diffusivity. In terms of non-linear systems analysis, this would be termed a \emph{parametric sensitivity analysis} being conducted by running an open loop of the dynamic system under a set of initial conditions and a subset of parameter values (here oxygen diffusivity).

When one wishes to be able to make conclusions about how the \emph{steady state} (or an equilibrium point) of the system is affected by the environmental driver, it is very important to note that the \emph{final state} (provided by the simulation) is not guaranteed to be a \emph{steady state}. The software does not provide the user with a steady state. In order to somewhat safely assume that the final state is a steady state, the user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappear, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper \citet{Limberger2023}, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation. Furthermore, the package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique.

Two numerical strategies for finding final states and their sensitivity to parameters are implemented. The first method runs a independent simulation for each combination of initial conditions and oxygen diffusivity (we term this the \emph{Replication method}). This is the method used in the \citet{Bush2017} study and was used to obtain the results in figures 3 and 4 of that article.

@@

The open source implementation and extension of the model used in \citet{Bush2017} provides the means of reproducing the results published while at the same time provides the means of making unique, innovative, and important investigations of how ecosystems change as the environment changes, and how biodiversity may modulate this.

The design of the package code and functionality is with reproducibility in mind (for other example of reproducibility directed software, see e.g. \citet{Correndo2023}, \citet{Spillner2022} or \citet{Fox2020}): the combination of all parameters being in a single parameter object and the return of the simulation as a result object which inherits the parameter values from the parameter object, provides a relatively easy-to-use framework to implement reproducible experiments.
